---
title: "Sturgeon Vertebrae Mechanics"
output: html_notebook
---
# Introduction
This notebook runs calculations and generates plot for the paper titled: Early tetrapod neck mobility modeled from an extant analogue (currently Nature manuscript 2025-12-33097).

Data from this notebook are represented in Supplementary Figures 3, 4, 5, Table 1, and Supplementary Tables 7, 8, and 9

Please email both dana.korneisel@ebc.uu.se and cdonatel@uw.edu with any questions or requests

# Libraries
The following libraries are required to run the code.
```{r, echo=FALSE,  message=FALSE}
# For reading in Data
library(readr)
library(readxl)

# For making plots
library(ggplot2)         
library(ggthemes)        
library(RColorBrewer)
library(cowplot)
library(scales)
library(ggnewscale)

# For filtering and sorting
library(dplyr)

# For tan_d function in mobility calculations
# Note that this is not presented in the paper, this was just for the authors interest
library(aspace)

```

# Reading in data
The following code reads in the material testing data used in the plots. 
```{r, echo=FALSE,  message=FALSE}
# This line reads in data from torsion trials
SturgeonTorsionOrig <- read_csv("SturgeonTorsion_2026.csv")

# This line reads in data from bending trials.
SturgeonBendingOrig <- read_csv("SturgeonBending_2026.csv")

```

# Processing Torsion Data - Polar Moment of Inertia
1) Map the position column to a factor. Since the positions are listed as numbers, R, by default, treats them as continuous numbers. Converting them to a factor lets us treat them as discrete values

2) Create two sub-data frames with the two different dissection conditions (whole vertebral column, "column") (notochord only, "notochord"). This filtering is required for torsion calculations, as the different cross-sections shapes and areas affect the polar moment of inertia (J, m^4). 

3) Calculate polar moment of inertia (J)
  3a) For the notochord dissection, estimate a circle (J = π*D^4/32) where D is the diameter of the notochord
  3b) For the vertebral column dissection, estimate a triangle (J = (b*h/144) x (4b^2 3h^2)) where h is the height of the estimated vertebral triangle (m) and b is the base of the estimated vertebral triangle
  - Note that all measurements are taken in mm, so we divide by 1000 to convert to meters

4) Merge data back into a single data frame: SturgeonTorsion
```{r}
# 1) Map position column to a factor
SturgeonTorsionOrig$Position <- as.factor(SturgeonTorsionOrig$Position)

# 2) Filter data by dissection condition.
STVert <- filter(SturgeonTorsionOrig, dissection == "column")
STNoto <- filter(SturgeonTorsionOrig, dissection == "notochord")

# 3a) Polar moment of inertia for notochord dissection
STNoto$J = (pi*(STNoto$diameter/1000)^4)/32

# 3b) Polar moment of inertia for vertebral column dissection
STVert$J = (((STVert$base/1000)*(STVert$height/1000))/144) * 
  (4*(STVert$base/1000)^2 + 3*(STVert$height/1000)^2)

# 4) Merge Data into single data frame
SturgeonTorsion <- rbind(STNoto, STVert)
```

# Processing Torsion Data - Torsional Mechanics
Note that all linear measurements are taken in mm, so we divide by 1000 to convert to meters

1) Calculate torsional rigidity (T*L/θ) where T is maximum torque (Nm), L is gage length (m), and θ is the angle at maximum torque (degrees). 

2) Calculate torsional stiffness (T/θ) where T is maximum torque (Nm) and θ is the angle at maximum torque (degrees).

3) Calculate torsional modulus (TL/θJ) where T is maximum torque (Nm), L is gage length (m), and θ is the angle at maximum torque (degrees), and J is the polar moment of inertia (m^4). Here, we simply divide by J, as we already calculated GJ in step 1.
```{r}
# 1) Torsional Rigidity
SturgeonTorsion$GJ <- (SturgeonTorsion$cycle1MaxLoad*(SturgeonTorsion$Gage/1000))/((SturgeonTorsion$cycle1MaxAngle*pi)/180)

# 2) Torsional Stiffness
SturgeonTorsion$GJL <- SturgeonTorsion$cycle1MaxLoad/((SturgeonTorsion$cycle1MaxAngle*pi)/180)

# 3) Torsion Modulus
SturgeonTorsion$G <- SturgeonTorsion$GJ/SturgeonTorsion$J
```

# Processing Bending Data - Second Moment of Inertia
1) Map the position column to a factor. Since the positions are listed as numbers, R, by default, treats them as continuous numbers. Converting them to a factor lets us treat them as discrete values

2) Create two sub-data frames with the two different dissection conditions (whole vertebral column, "column") (notochord only, "notochord"). This filtering is required for torsion calculations, as the different cross-sections shapes and areas affect the second moment of area (I, m^4). 

3) Calculate second moment of area (I, m^4) for notochord dissection
For the notochord dissection, estimate a circle (I = π*r^4/32) where r is the radius of the notochord. Note that diameter measurements were taken in mm, so we divide by 1000 to convert to meters

4) Create two sub-data frames with the two different bending orientations (dorsal and lateral). This is because second moment of area is dependent on the axis of bending, and it is not symmetrical for a triangle.

5) Calculate second moment of inertia (I, m^4) for vertebral column dissection
  5a) For the lateral orientation (I = h*b^3/48) where h is the height of the estimated vertebral triangle (m) and b is the base of the estimated vertebral triangle
  5b) For the dorsal orientation (I = h^3*b/48) where h is the height of the estimated vertebral triangle (m) and b is the base of the estimated vertebral triangle
  - Note that base and height measurements were taken in cm, so we divide by 100 to convert to meters

6) Merge data back into a single data frame: SturgeonBending
```{r}
# 1) Map position column to a factor
SturgeonBendingOrig$Position <- as.factor(SturgeonBendingOrig$Position)

# 2) Filter data by dissection condition.
SBVert <- filter(SturgeonBendingOrig, dissection == "column")
SBNoto <- filter(SturgeonBendingOrig, dissection == "notochord")

# 3) Second moment of area for notochord dissection
SBNoto$I = (pi*(SBNoto$diameter/1000/2)^4)/4

# 4) Filter by bending orientation
SBVertLat = filter(SBVert, orientation == "Lateral")
SBVertDVT = filter(SBVert, orientation == "Dorso")

# 5a) Calculate I for Lateral Bending
SBVertLat$I <- (SBVertLat$height/100 * (SBVertLat$base/100)^3)/48 

# 5b) Calculate I for Dorsal Bending
SBVertDVT$I <- ((SBVertDVT$height/100)^3 * SBVertDVT$base/100)/48 

# 6) Merge back into a single data frame
SBVert2 <- rbind(SBVertLat, SBVertDVT)
SturgeonBending <- rbind(SBNoto, SBVert2)
```

# Processing Bending Data - Bending Mechanics
1) Calculate flexural rigidity (rF/d) where rF represents the bending moment (M). F is the maximum force (N), r is the 3-point bending rig span (m), and d is the maximum displacement (m). Note that displacement and span was taken in mm so we divide by 100 to convert to meters, and the displacement was taken in mm so we divide by 1000 to convert to meters. 

2) Calculate bending stiffness (F/d) where F is the maximum force (N) and d is the maximum displacement (m). Note that displacement was taken in mm, so we divide by 1000 to convert to meters

3) Calculate Youngs modulus (F*L^3/48dI) where F is the maximum force (N), L is the 3-point bending rig span (m), and d is the maximum displacement (m). Here, we simply divide by I, as we already calculated EI in step 1. 
```{r}
# 1) Flexural Rigidity (rF/d)
SturgeonBending$EI <- (SturgeonBending$cycle1MaxLoad*(SturgeonBending$length/100))/(SturgeonBending$MaxBend/1000)

SturgeonBending$EIold <- (SturgeonBending$cycle1MaxLoad*(SturgeonBending$length/1000))/(SturgeonBending$MaxBend/1000)

# 2) Bending Stiffness (F/d)
SturgeonBending$EL <- SturgeonBending$cycle1MaxLoad/(SturgeonBending$MaxBend/1000)

# 3) Yonung's Modulus (E)
SturgeonBending$E <- SturgeonBending$EI/SturgeonBending$I
```

# Supplemental Figures
First, separate data by bending orientation for plotting
```{r}
SBDVT = filter(SturgeonBending, orientation == "Dorso")
SBLat = filter(SturgeonBending, orientation == "Lateral")
```

The code below creates a plot showing Youngs modulus (E) and Flexural Rigidity (EI) for dorsoventral bending separated by position and dissection condition. Un-comment out the ggsave() line to save the plot in your working directory.
```{r}
# Make the two plots, but remove legend from the second one
p1 <- ggplot(SBDVT, aes(x = dissection, y = E/1e9, fill = Position)) +
  geom_boxplot() +
  geom_jitter(aes(color = Position), alpha = 0.4, size = 2, show.legend = FALSE,
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75)) +
  scale_fill_brewer(palette = "Blues") +
  scale_color_brewer(palette = "Blues") +
  xlab(NULL) +
  ylab("Flexural Modulus (E, GPa)") +
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) + 
  theme(
    text = element_text(size = 12),
    legend.position = "top",
    legend.spacing.x = unit(0.2, "cm"),  # reduce horizontal spacing
    legend.spacing.y = unit(0.1, "cm"),  # reduce vertical spacing
    legend.key.size = unit(0.5, "cm"))   # shrink legend keys (boxes)

p2 <- ggplot(SBDVT, aes(x = dissection, y = EI, fill = Position)) +
  geom_boxplot() +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    alpha = 0.4, size = 2, show.legend = FALSE, color = "grey") +
  scale_fill_brewer(palette = "Blues") +
  xlab("Dissection Condition") +
  ylab(expression("Flexural Rigidity (EI, N·m"^2*")"))+
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) + 
  theme(
    text = element_text(size = 12),
    legend.position = "none")   # remove legend

# Align them vertically with cowplot
DV <- plot_grid(p1, p2, ncol = 1, align = "v", axis = "lr", rel_heights = c(1,0.9))
DV
#ggsave("dorsoVentralBending.png",plot = DV, width = 4, height = 6, units = "in")
```

New Plots of Youngs modulus (E) and Flexural Rigidity (EI) for dorsoventral bending showing the 
raw datapoints jittered so that you can see the
```{r}
SBDVT$box_alpha <- ifelse(SBDVT$Position == "Position 4", 0, 1)
SBDVT$box_line  <- ifelse(SBDVT$Position == "Position 4", NA, "black")

# This will create a variable to remove the box and mean from the plots if there
# are fewer than three data points. In this case only for the notocord at Position 4
SBDVT <- SBDVT %>%
  group_by(dissection, Position) %>%
  mutate(n_group   = n(),
         box_alpha = ifelse(n_group < 3, 0, 1),   # <3 points = no real box
         box_line  = ifelse(n_group < 3, NA, "black")
         ) %>%
  ungroup()

p1 <- ggplot(SBDVT, aes(x = dissection, y = E/1e9)) +
  geom_boxplot(aes(alpha = box_alpha, colour = box_line, group = interaction(dissection, Position)),
    fill = "grey85", outlier.shape = 3, outlier.color = "darkred",outlier.size = 2.5,
    position = position_dodge2(preserve = "single")) +
  scale_alpha_identity() +
  scale_colour_identity() +
  geom_jitter(aes(fill = Position),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    shape = 21, colour = "black",stroke = 0.3, size = 2) +
  scale_fill_brewer(palette = "Blues") +
  xlab(NULL) +
  ylab("Flexural Modulus (E, GPa)") +
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) + 
  theme(
    text = element_text(size = 12),
    legend.position = "top",
    legend.spacing.x = unit(0.1, "cm"),
    legend.spacing.y = unit(0.1, "cm"),
    legend.key.size = unit(0.25, "cm"))+
  annotate("text", x = 1.28, y = max(SBDVT$E)/1e9+15, label = "n < 3", size = 3.5)

p2 <- ggplot(SBDVT, aes(x = dissection, y = EI)) +
  geom_boxplot(aes(alpha = box_alpha, colour = box_line, group = interaction(dissection, Position)),
    fill = "grey85", outlier.shape = 3, outlier.color = "darkred",outlier.size = 2.5,
    position = position_dodge2(preserve = "single")) +
  scale_alpha_identity() +
  scale_colour_identity() +
  geom_jitter(aes(fill = Position),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    shape = 21, colour = "black", stroke = 0.3, size = 2) +
  scale_fill_brewer(palette = "Blues") +
  xlab("Dissection Condition") +
  ylab(expression("Flexural Rigidity (EI, N·m"^2*")")) +
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) + 
  theme(text = element_text(size = 12), legend.position = "none") +
  annotate("text", x = 1.28, y = max(SBDVT$EI)+1, label = "n < 3", size = 3.5)

# Align them vertically with cowplot
DV <- plot_grid(p1, p2, ncol = 1, align = "v", axis = "lr", rel_heights = c(1,0.9))
DV
#ggsave("dorsoVentralBending02.png",plot = DV, width = 5, height = 6, units = "in")
```

The code below creates a plot showing Youngs modulus (E) and Flexural Rigidity (EI) for lateral bending separated by position and dissection condition. Un-comment out the ggsave() line to save the plot in your working directory.
```{r}
# Make the two plots, but remove legend from the second one
p1 <- ggplot(SBLat, aes(x = dissection, y = E/1e9, fill = Position)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Blues") +
  xlab(NULL) +
  ylab("Flexural Modulus (E, GPa)") +
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) + 
  theme(
    text = element_text(size = 12),
    legend.spacing.x = unit(0.01, "cm"),  # reduce horizontal spacing
    legend.spacing.y = unit(0.1, "cm"),  # reduce vertical spacing
    legend.key.size = unit(0.4, "cm"),   # shrink legend keys (boxes)
    legend.key.width = unit(0.3, "cm"), # shrink box width
    legend.position = "top")+
  guides(fill = guide_legend(nrow = 1)) 

p2 <- ggplot(SBLat, aes(x = dissection, y = EI, fill = Position)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Blues") +
  xlab("Dissection Condition") +
  ylab(expression("Flexural Rigidity (EI, N·m"^2*")"))+
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) + 
  theme(
    text = element_text(size = 12),
    legend.position = "none")   # remove legend

# Align them vertically with cowplot
Lat <- plot_grid(p1, p2, ncol = 1, align = "v", axis = "lr", rel_heights = c(1,0.9))
Lat
#ggsave("lateralBending.png",plot = Lat, width = 4, height = 6, units = "in")
```

New Plots of Youngs modulus (E) and Flexural Rigidity (EI) for lateral bending showing the 
raw datapoints jittered so that you can see them
```{r}
SBLat$box_alpha <- ifelse(SBLat$Position == "Position 4", 0, 1)
SBLat$box_line  <- ifelse(SBLat$Position == "Position 4", NA, "black")

# This will create a variable to remove the box and mean from the plots if there
# are fewer than three data points. In this case only for the notocord at Position 8
SBLat <- SBLat %>%
  group_by(dissection, Position) %>%
  mutate(n_group   = n(),
         box_alpha = ifelse(n_group < 3, 0, 1),   # <3 points = no real box
         box_line  = ifelse(n_group < 3, NA, "black")
         ) %>%
  ungroup()

ann1 <- data.frame(dissection = "column", x = 8, y = max(SBLat$E/1e9, na.rm = TRUE)+1, label = "n < 3")

p1 <- ggplot(SBLat, aes(x = Position, y = E/1e9)) +
  geom_boxplot(aes(alpha = box_alpha, colour = box_line),
    fill = "grey85", position = position_dodge2(preserve = "single"),
    outlier.shape = 3, outlier.color = "darkred",outlier.size = 2.5) +
  scale_alpha_identity() +
  scale_colour_identity() +
  geom_jitter(aes(fill = Position),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    shape = 21, colour = "black", stroke = 0.3, size = 2) +
  geom_text(data = ann1, aes(x = x, y = y, label = label), size = 3.5, inherit.aes = FALSE) +
  scale_fill_brewer(palette = "Blues") +
  facet_wrap(~ dissection, scales = "free_x", strip.position = "bottom") +
  xlab(NULL) +
  ylab("Flexural Modulus (E, GPa)") +
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) + 
  theme(
    text = element_text(size = 12),
    legend.position = "top",
    legend.spacing.x = unit(0.1, "cm"),
    legend.spacing.y = unit(0.01, "cm"),
    legend.key.size = unit(0.2, "cm"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.background = element_blank(),
    strip.placement = "outside") 

ann2 <- data.frame(dissection = "column", x = 8, y = max(SBLat$EI, na.rm = TRUE)+1, label = "n < 3")

p2 <- ggplot(SBLat, aes(x = Position, y = EI)) +
  geom_boxplot(aes(alpha = box_alpha, colour = box_line),
    fill = "grey85", position = position_dodge2(preserve = "single"),
    outlier.shape = 3, outlier.color = "darkred",outlier.size = 2.5) +
  scale_alpha_identity() +
  scale_colour_identity() +
  geom_jitter(aes(fill = Position),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    shape = 21, colour = "black", stroke = 0.3, size = 2) +
  geom_text(data = ann2, aes(x = x, y = y, label = label), size = 3.5, inherit.aes = FALSE) +
  scale_fill_brewer(palette = "Blues") +
  facet_wrap(~ dissection, scales = "free_x", strip.position = "bottom") +
  xlab("Dissection Condition") +
  ylab(expression("Flexural Rigidity (EI, N·m"^2*")")) +
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) + 
  theme(
    text = element_text(size = 12),
    legend.position = "none",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.background = element_blank(),
    strip.placement = "outside")

# Align them vertically with cowplot
Lat <- plot_grid(p1, p2, ncol = 1, align = "v", axis = "lr", rel_heights = c(1,0.85))
Lat
#ggsave("lateralBending02.png",plot = Lat, width = 5, height = 6, units = "in")
```

The code below creates a plot showing Torsional modulus (G), Torsional Rigidity (GI), and Torsional Stiffness for lateral bending separated by position and dissection condition. Un-comment out the ggsave() line to save the plot in your working directory.
# Updated to include jittered points to match the aesthetic for the bending plots
```{r}
# Make the two plots, but remove legend from the second one
p2 <- ggplot(SturgeonTorsion, aes(x = dissection, y = GJ)) +
  geom_boxplot(aes(group = interaction(dissection, Position)),
    fill = "grey85", outlier.shape = 3, outlier.color = "darkred",outlier.size = 2.5,
    position = position_dodge2(preserve = "single")) +
  geom_jitter(aes(fill = Position),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    shape = 21, colour = "black", stroke = 0.3, size = 2) +
  scale_fill_brewer(palette = "Blues") +
  xlab(NULL) +
  ylab(expression("Torsional Rigidity (GJ, N·m"^2*")"))+
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.0001)) + 
  theme(text = element_text(size = 12),
        legend.position = "none")   # remove legend
  

p1 <- ggplot(SturgeonTorsion, aes(x = dissection, y = G/1e9, fill = Position)) +
  geom_boxplot(aes(group = interaction(dissection, Position)),
    fill = "grey85", outlier.shape = 3, outlier.color = "darkred",outlier.size = 2.5,
    position = position_dodge2(preserve = "single")) +
  geom_jitter(aes(fill = Position),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    shape = 21, colour = "black", stroke = 0.3, size = 2) +
  scale_fill_brewer(palette = "Blues") +
  xlab(NULL) +
  ylab("Torsional Modulus (G, GPa)") +
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.0001)) + 
  theme(text = element_text(size = 12),
        legend.position = "top")

p3 <- ggplot(SturgeonTorsion, aes(x = dissection, y = GJL, fill = Position)) +
  geom_boxplot(aes(group = interaction(dissection, Position)),
    fill = "grey85", outlier.shape = 3, outlier.color = "darkred",outlier.size = 2.5,
    position = position_dodge2(preserve = "single")) +
  geom_jitter(aes(fill = Position),
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.75),
    shape = 21, colour = "black", stroke = 0.3, size = 2) +
  scale_fill_brewer(palette = "Blues") +
  xlab("Dissection Condition") +
  ylab("Torsional Stiffness (GJ/L, Nm)") +
  theme_classic() +
  scale_y_continuous(labels = number_format(accuracy = 0.0001)) + 
  theme(text = element_text(size = 12),
        legend.position = "none")

# Align them vertically with cowplot
Tor <- plot_grid(p1, p2, p3, ncol = 1, align = "v", axis = "lr", rel_heights = c(1,0.9,0.9))
Tor
#ggsave("torsion02.png",plot = Tor, width = 5, height = 8, units = "in")
```

# Summary Statistics Function
This function creates summary statistics for different groups of data. Please add the function to your workspace by running this chunk before running the "Summary Statistics" section below.
```{r}
summarize_by_groups <- function(df, cont_var, group1, group2) {
  df %>%
    group_by(across(all_of(c(group1, group2)))) %>%
    summarise(
      mean = mean(.data[[cont_var]], na.rm = TRUE),
      median = median(.data[[cont_var]], na.rm = TRUE),
      sd = sd(.data[[cont_var]], na.rm = TRUE),
      n = n(),
      min = min(.data[[cont_var]], na.rm = TRUE),
      max = max(.data[[cont_var]], na.rm = TRUE),
      .groups = "drop"
    )
}
```

# Summary Statistics
This section uses the function above to calculate means, medians, and standard deviations by dissection condition and position for the mechanical properties calculated.
```{r}
bendLateral = filter(SturgeonBending, orientation == "Lateral")
bendDorso = filter(SturgeonBending, orientation == "Dorso")

summarize_by_groups(bendLateral, "E", "dissection", "Position")
summarize_by_groups(bendLateral, "EI", "dissection", "Position")
summarize_by_groups(bendLateral, "cycle1MaxLoad", "dissection", "Position")
summarize_by_groups(bendLateral, "cycle1Work", "dissection", "Position")

summarize_by_groups(bendDorso, "E", "dissection", "Position")
summarize_by_groups(bendDorso, "EI", "dissection", "Position")
summarize_by_groups(bendDorso, "cycle1MaxLoad", "dissection", "Position")
summarize_by_groups(bendDorso, "cycle1Work", "dissection", "Position")

summarize_by_groups(SturgeonTorsion, "G", "dissection", "Position")
summarize_by_groups(SturgeonTorsion, "GJ", "dissection", "Position")
summarize_by_groups(SturgeonTorsion, "GJL", "dissection", "Position")
summarize_by_groups(SturgeonTorsion, "cycle1MaxLoad", "dissection", "Position")
```

# Exploratory Plots of Mechanical Testing Data
The plots below are not included in the paper. They serve only to explore the data plotted in different ways.
```{r}
ggplot(SturgeonTorsion, aes(x = dissection, y = GJ, fill = Position))+
  geom_boxplot()+
  scale_fill_brewer(palette = "Reds")+
  xlab("Dissection Condition")+
  ylab("Torsional Rigidity (GJ, Nm^2/rad)")+
  theme_minimal()+
  theme(text = element_text(size = 20))

ggplot(SturgeonTorsion, aes(x = dissection, y = G/10^9, fill = Position))+
  geom_boxplot()+
  scale_fill_brewer(palette = "Reds")+
  xlab("Dissection Condition")+
  ylab("Torsion Modulus (G, GPa)")+
  theme_minimal()+
  theme(text = element_text(size = 20))

## Removed outliers to make things clearer
## To add them back, remove the ylim line and the + sign before it
ggplot(SturgeonBending, aes(x = dissection, y = EI, fill = Position))+
  geom_boxplot()+
  scale_fill_brewer(palette = "Blues")+
  xlab("Dissection Condition")+
  ylab("Flexural Rigidity (EI, Nm^2)")+
  theme_minimal()+
  theme(text = element_text(size = 20))+
  ylim(0, 1.5)

ggplot(SturgeonBending, aes(x = dissection, y = E/10^9, fill = Position))+
  geom_boxplot()+
  scale_fill_brewer(palette = "Blues")+
  xlab("Dissection Condition")+
  ylab("Flexural Modulus (E, GPa)")+
  theme_minimal()+
  theme(text = element_text(size = 20))

SBLat = filter(SturgeonBending, orientation == "Lateral")
SBLat$Position <- factor(
  SBLat$Position,
  levels = c("Position 1", "Position 2", "Position 3", "Position 4", 
             "Position 5", "Position 6", "Position 7", "Position 8"),
  labels = c("1", "2", "3", "4", "5", "6", "7", "8"))

SBDVT = filter(SturgeonBending, orientation == "Dorso")
SBDVT$Position <- factor(
  SBDVT$Position,
  levels = c("Position 1", "Position 2", "Position 3", "Position 4"),
  labels = c("1", "2", "3", "4"))

ggplot(SBLat, aes(x = dissection, y = EI, fill = Position))+
  geom_boxplot()+
  scale_fill_brewer(palette = "Blues")+
  xlab("Dissection Condition")+
  ylab("Flexural Rigidity (EI, Nm^2)")+
  theme_minimal()+
  theme(text = element_text(size = 20))
```

## Exploratory Mobility Plots
The data manipulation and plots below are not included in the paper. They serve only to explore the mobility estimates performed in excel graphically. 
```{r}
mobility <- read_excel("Taxaformobilityestimates.xlsx",sheet = "forR")

mobility$latColF = ( (tan_d(mobility$LatBendDegrees)*mobility$Length.m)*
                   3*mobility$E.Pa.LatCol*mobility$IlatCol ) /
                  (mobility$Length.m^3)
mobility$LatColForceReq = mobility$MaxMuscleArea/10000*200000

mobility$DVColF = ( (tan_d(mobility$DVBendDegrees)*mobility$Length.m)*
                   3*mobility$E.Pa.Dcol*mobility$IDVCol ) /
                  (mobility$Length.m^3)

mobility$TColT = (as_radians(mobility$TorsionDegrees)*mobility$G.Pa.Col*mobility$Jcol)/mobility$Length.m

mobility$latNotF = ( (tan_d(mobility$LatBendDegrees)*mobility$Length.m)*
                   3*mobility$E.Pa.LatNot*mobility$Inoto ) /
                  (mobility$Length.m^3)

mobility$DVNotF = ( (tan_d(mobility$DVBendDegrees)*mobility$Length.m)*
                   3*mobility$E.Pa.Dnot*mobility$Inoto ) /
                  (mobility$Length.m^3)

mobility$TNotT = (as_radians(mobility$TorsionDegrees)*mobility$G.Pa.Not*mobility$Jnoto)/mobility$Length.m

hline_data <- mobility %>%
  distinct(Taxon, LatColForceReq)

ggplot(mobility, aes(x = LatBendDegrees, y = latColF, color = Taxon)) +
  geom_point() +
  geom_line()+
  geom_hline(data = hline_data, 
             aes(yintercept = LatColForceReq, color = Taxon), linewidth = 1) +
  theme_classic()+
  ylim(0, 30000)
```
