---
title: "Sturgeon Material Testing Data Extractor"
output: html_notebook
---
##NOTE##
To run the code below you must add a function to your workspace. For the 
sturgeon project, that's either a function to extract torsional data or to 
extract three point bending data. Both types of functions are available in the
https://github.com/CDonatelli/Korneisel_TetNecks_2026 repository and are 
derived from function in the https://github.com/CDonatelli/CrshR repository
for general material testing data extraction of biological and biologically 
inspired samples. 

Please contact Cassandra Donatelli (cdonatel@uw.edu) with any questions or 
requests for expansions of the scripts

#Required Libraries
```{r}
library(readr)
library(ggplot2)
library(pracma)
library(tcltk)
library(TTR)
```

### 3PT BENDING DATA PROCESSING ###

#Enter in your test data
This includes the number of cycles, maximum deflection (mm) of the upper roller 
(programmed by the user), and span between the two support rollers (mm)
```{r}
# This is the same for all bending trials
BendCycles = 5

# Maximum deflection (mm)
maxExtLat = 10
maxExtDor = 5

##################
# NOTE: This changes for different trials

# See the data table for more information
Length = 20 # Gage length (mm)

#################
```

### Add the bend_3pt_cyclic_TimeForce Function into your workspace ###

# Read in data and call the function
This is for Lateral Bends
```{r}
# Select and read in the data
dataFile = file.choose()
data <- read.csv(dataFile, skip = 3, header = F)
colnames(data)<- c("Time","Load")
  
### This code finds the timing of each cycle
dataPts = nrow(data)
cycleLength = dataPts/bendCycles
userPts = floor(ppoints(bendCycles+1,1)*dataPts)
userPts[1]=1
userPts = data$Time[userPts]

# This plots it to make sure you are good to go
# The red lines should be seperating the start and stop of each cycle
ggplot(data, aes(x = Time, y = Load)) +
  geom_line() +
  geom_vline(xintercept = userPts, color = "red", linetype = "dashed") +
  labs(title = "Load over Time with Cycle Points Marked",
       x = "Time", y = "Load") +
  theme_minimal()
```

# Run the function
```{r}
dataOut <- bend_3pt_cyclic_TimeDisp(data, Length, bendCycles, userPts, maxExtLat)

# Display Results so that it's easy to copy into table
results <- dataOut$results
knitr::kable(result)

# Display the plot
dataOut$plot
```

### TORSION DATA ###

#Enter in your test data
This includes the number of cycles, maximum deflection (mm) of the upper roller 
(programmed by the user), and span between the two support rollers (mm)
```{r}
# This is the same for all bending trials
torsionCycles = 3

# Maximum Angle (degrees)
maxAngle = 30

##################
# NOTE: This changes for different trials

# See the data table for more information
Length = 50 # Gage length (mm)

#################
```

### Add the torsion_cyclic_TimeForce Function into your workspace ###

# Read in data and call the function
This is for Lateral Bends
```{r}
# Select and read in the data
dataFile = file.choose()
data <- read.csv(dataFile, skip = 3, header = F)
colnames(data)<- c("Time","Load", "Angle")

# Remove the extra hold time at the end so we can easily crop cycles
data <- data[data$Time <= (max(data$Time) - 1.5), ]
  
### This code finds the timing of each cycle
dataPts = nrow(data)
cycleLength = dataPts/torsionCycles
userPts = floor(ppoints(torsionCycles+1,1)*dataPts)
userPts[1]=1
userPts = data$Time[userPts]

# This plots it to make sure you are good to go
# The red lines should be seperating the start and stop of each cycle
ggplot(data, aes(x = Time, y = Load)) +
  geom_line() +
  geom_vline(xintercept = userPts, color = "red", linetype = "dashed") +
  labs(title = "Load over Time with Cycle Points Marked",
       x = "Time", y = "Load") +
  theme_minimal()
```

# Run the function
```{r}
dataOut <- torsion_cyclic_TimeForce(data, Length, torsionCycles, userPts)

# Display Results so that it's easy to copy into table
results <- dataOut$results
knitr::kable(results)

# Display the plot
dataOut$plot
```
